---
title: "essentiality enrichment"
output: html_document
date: "2024-07-12"
---

```{r}
library(data.table)
library(tidyr)
library(ggplot2)
library(ggthemes)
library(tidyverse)
library(readxl)
library(dplyr)
library(stringr)
```

```{r}
# read and extract vector of essential and non-essential genes
essential_input <- read_excel("Z:/time_course_BDQ/NCBI_SRA/essentiality_analysis/essential_genes_Zhang_2012.xlsx")
#pull essential genes, both standard name and Rv Number and pool them
essential_input <- essential_input %>% mutate(Gene = ifelse(Gene == "-", `Rv Number`, Gene))
essential_genes <- essential_input %>%
  filter(Call %in% c("E", "D")) %>%
  select(Gene)
#essential_rv_numbers <- essential_input %>%
  #filter(Call %in% c("E", "D")) %>%
  #select(`Rv Number`)
#names(essential_rv_numbers)[names(essential_rv_numbers) == "Rv Number"] <- "Gene"
#essential_combined <- rbind(essential_genes, essential_rv_numbers)
#now the same thing for non-essential genes
non_essential_genes <- essential_input %>%
  filter(Call %in% c("NE")) %>%
  select(Gene)
#non_essential_rv_numbers <- essential_input %>%
  #filter(Call %in% c("NE")) %>%
  #select(`Rv Number`)
#names(non_essential_rv_numbers)[names(non_essential_rv_numbers) == "Rv Number"] <- "Gene"
#non_essential_combined <- rbind(non_essential_genes, non_essential_rv_numbers)
included_genes <- c(essential_genes, non_essential_genes)  # uncertain genes and genes not in the essential genes table are excluded from analysis
included_genes<- unlist(included_genes)
#load up merged SV data (annotated), as well as all the isolate names
structural_variants <- fread("Z:/time_course_BDQ/NCBI_SRA/essentiality_analysis/merged_annotated_final.csv")
svs2<- read_tsv("Z:/time_course_BDQ/NCBI_SRA/essentiality_analysis/output.txt")
filtered_svs2 <- subset(svs2,SVLen >= -100000 & SVLen <= 100000)
#keep only coding sequence variants
#structural_variants <- structural_variants %>% 
  #filter(str_detect(Consequence, "coding_sequence_variant"))
pacbio_isolates <- fread("Z:/assembly/pacbio_sra_complete_final.txt", header = F)
nanopore_isolates <- fread("Z:/assembly/nanopore_sra_complete_final.txt", header = F)
isolates <- rbind(pacbio_isolates,nanopore_isolates)
isolates <- isolates$V1
```
```{r}
# Extract structural variants and convert table to long form
#svs_table <- structural_variants[grep("SV", Consequence)]
svs_table <- data.table(gather(structural_variants, isolate, presence, 14:291))
svs_table <- svs_table[presence == 1] 
#included_genes <- list(essential_combined = essential_combined, non_essential_combined = non_essential_combined)
essential_genes <- unlist(essential_genes)
non_essential_genes <- unlist(non_essential_genes)
svs_essential <- svs_table[Symbol %in% essential_genes, c("Symbol", "isolate")] 
svs_non_essential <- svs_table[Symbol %in% non_essential_genes, c("Symbol", "isolate")]
svs_all<-rbind(svs_essential,svs_non_essential)
colnames(svs_all) <- c("Gene","isolate")
svs_all$Gene_Group <- "error"
svs_all[Gene %in% essential_genes]$Gene_Group <- "Essential"
svs_all[Gene %in% non_essential_genes]$Gene_Group <- "Non_Essential"# Use long form table to count for each gene the number of isolates with at least one structural variant in that gene 
#generate file that counts essential vs non essential svs by isolate 
setkey(svs_all, isolate, Gene_Group)
svs_by_iso <- svs_all[CJ(isolate, Gene_Group, unique = T), .N, by=.EACHI]
colnames(svs_by_iso) <- c("Isolate", "Gene_Group", "Count")
```
```{r}
# use long form table to count for each gene the number of isolates with at least one SV in that gene
svs_by_gene <- data.table(table(svs_table$Symbol)) # number of times each gene appears in fs_cons = the number of isolates with at least one frameshift in that gene (now that we've removed the duplicated isolate gene pairs)
colnames(svs_by_gene) <- c("Gene", "Count") 
no_svs_genes <- data.table(Count = 0, Gene = setdiff(included_genes, svs_by_gene$Gene))# the genes not in fs_cons have an SV in 0 isolates. 
svs_by_gene <- rbind(svs_by_gene, no_svs_genes)

```
```{r}
#test whether SVs are more frequent in non-essential genes
t.test(svs_by_gene[Gene %in% essential_genes]$Count, svs_by_gene[Gene %in% non_essential_genes]$Count)
median(svs_by_gene[Gene %in% essential_genes]$Count)
median(svs_by_gene[Gene %in% non_essential_genes]$Count)
nrow(svs_by_gene[(Count>=100)&(Gene %in% essential_genes)])
nrow(svs_by_gene[(Count>=100)&(Gene %in% non_essential_genes)])
wilcox.test(svs_by_gene[Gene %in% essential_genes]$Count, svs_by_gene[Gene %in% non_essential_genes]$Count)

```
```{r}
# Graphs comparing LOF rates between Essential and Non-essential Genes (isolate counts by gene)
svs_by_gene$Group <- "Error"
svs_by_gene[Gene %in% essential_genes]$Group <- "Essential"
svs_by_gene[Gene %in% non_essential_genes]$Group <- "Nonessential"
p <- ggplot(svs_by_gene, aes(x=Count, color=Group))
p <- p + geom_histogram(binwidth = 1)
p <- p + xlab("Structural Variant Frequency (Isolate Count)") + ylab("Genes")
p 
```
```{r}
ggplot(svs_by_gene[Gene %in% essential_genes],aes(x=Count))+geom_histogram(binwidth=1)+xlim(0,20)
ggplot(svs_by_gene[Gene %in% non_essential_genes],aes(x=Count))+geom_histogram(binwidth=1)+xlim(0,20)
```
p <- ggplot(svs_by_gene[Gene %in% essential_genes]$Count], aes(x=Count, Fill=Group))
p <- p + geom_histogram(binwidth = 1) + theme_tufte(base_size = 16)
p <- p + xlab("Frameshift Frequency in Nonessential Genes (Isolate Count)") + ylab("Genes")
p 
p <- ggplot(fs_cons_by_gene[Group=="Essential"], aes(x=Count, Fill=Group))
p <- p + geom_histogram(binwidth = 1) + theme_tufte(base_size = 16)
p <- p + xlab("Frameshift Frequency in Essential Genes (Isolate Count)") + ylab("Genes")
p 
p <- ggplot(fs_cons_by_gene[Group %in% c("Essential", "Other_Nonessential")], aes(y=Count, x=Group))
p <- p + geom_boxplot()
p <- p + xlab("Frameshift Frequency (Isolate Count)") + theme_tufte(base_size = 24)
p 
#ggsave("Y:/workspace/nrd/gcdd-sweden/new-repurposed-drug-resistance/data/pre_prog_use/enrichment/fs_ess_isocount.png", width = 15, height = 12, units = 'in', dpi = 300)
#ggsave("Y:/workspace/nrd/gcdd-sweden/new-repurposed-drug-resistance/data/pre_prog_use/enrichment/fs_ess_isocount.svg")


```